Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-03-17 paper.qmd
Updated the model and proj for pbmc best

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
#| label: hexbin-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 60
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "pc"
)

df_bin_centroids_pbmc <- pbmc_model$df_bin_centroids
df_bin_pbmc <- pbmc_model$df_bin

## Triangulate bin centroids
tr1_object_pbmc <- tri_bin_centroids(
  df_bin_centroids_pbmc, x = "c_x", y = "c_y")
tr_from_to_df_pbmc <- gen_edges(
  tri_object = tr1_object_pbmc)

## Compute 2D distances
distance_pbmc <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_pbmc,
  start_x = "x_from",
  start_y = "y_from",
  end_x = "x_to",
  end_y = "y_to",
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_pbmc <- find_lg_benchmark(
  distance_edges = distance_pbmc,
  distance_col = "distance")

tr_df <- distinct(tibble(
  x = c(tr_from_to_df_pbmc[["x_from"]], tr_from_to_df_pbmc[["x_to"]]), 
  y = c(tr_from_to_df_pbmc[["y_from"]], tr_from_to_df_pbmc[["y_to"]])))

@@

# Combine with the true model for visualization
df <- dplyr::bind_rows(scaled_pbmc_data_model |> mutate(type = "model"),
                       scaled_pbmc_data |> mutate(type = "data"))

@@

<!--Fit the best model and compute error-->
```{r}
#| label: num-bins-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 30
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled_best,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "pc"
)

df_bin_centroids_pbmc <- pbmc_model$df_bin_centroids
df_bin_pbmc <- pbmc_model$df_bin

## Triangulate bin centroids
tr1_object_pbmc <- tri_bin_centroids(
  df_bin_centroids_pbmc, x = "c_x", y = "c_y")
tr_from_to_df_pbmc <- gen_edges(
  tri_object = tr1_object_pbmc)

## Compute 2D distances
distance_pbmc <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_pbmc,
  start_x = "x_from",
  start_y = "y_from",
  end_x = "x_to",
  end_y = "y_to",
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_pbmc <- find_lg_benchmark(
  distance_edges = distance_pbmc,
  distance_col = "distance")

tr_df <- distinct(tibble::tibble(
  x = c(tr_from_to_df_pbmc[["x_from"]], tr_from_to_df_pbmc[["x_to"]]),
  y = c(tr_from_to_df_pbmc[["y_from"]], tr_from_to_df_pbmc[["y_to"]])))

@@

trimesh_removed_pbmc_best <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_pbmc_scaled_best,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.3,
    color = "#999999"
  )  +
  interior_annotation("b1", cex = 2) +
  theme(
    aspect.ratio = 1
  )

## Compute error
error_df <- augment(
  df_bin_centroids = df_bin_centroids_pbmc,
  df_bin = df_bin_pbmc,
  training_data = training_data_pbmc,
  newdata = NULL,
  type_NLDR = "UMAP",
  col_start = "pc")

## Categorize error

error_df <- error_df |>
  mutate(type = case_when(
    row_wise_abs_error <= 5 ~ "error 0-5",
    row_wise_abs_error <= 10 ~ "error 5-10",
    row_wise_abs_error <= 15 ~ "error 10-15",
    row_wise_abs_error <= 20 ~ "error 15-20",
    row_wise_abs_error <= 25 ~ "error 20-25",
    .default = "error greter than 25"
  )) |>
  mutate(type = factor(type, levels = c(
    "error 0-5", "error 5-10", "error 10-15", "error 15-20", "error 20-25", 
    "error greter than 25")))

## To join embedding
error_df <- error_df |>
  bind_cols(umap_pbmc_scaled_best |>
              select(-ID))

error_plot_pbmc_best <- error_df |>
  ggplot(aes(x = UMAP1,
             y = UMAP2,
             color = type,
             group = ID)) +
  geom_point(alpha=0.5,
             size = 0.1) +
  interior_annotation("b", sc_ltr_pos)

@@

# Combine with the true model for visualization
df <- dplyr::bind_rows(scaled_pbmc_data_model |> mutate(type = "model"),
                       scaled_pbmc_data |> mutate(type = "data"))

@@

projection_scaled <- projection * 1.23

projected <- as.matrix(scaled_pbmc_data) %*% projection_scaled

projected_df <- projected |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  #dplyr::mutate(type = df_exe$type) |>
  dplyr::mutate(ID = dplyr::row_number())

projected_model <- as.matrix(scaled_pbmc_data_model) %*% projection_scaled

model_df <- dplyr::left_join(
  distance_df_small_edges_pbmc |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

axes_obj <- gen_axes(
  proj = projection,
  limits = 1,
  axis_pos_x = -0.48,
  axis_pos_y = -0.48,
  axis_labels = names(scaled_pbmc_data),
  threshold = 0.082)

@@

projection_scaled <- projection * 1.25

projected <- as.matrix(scaled_pbmc_data) %*% projection_scaled

projected_df <- projected |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  #dplyr::mutate(type = df_exe$type) |>
  dplyr::mutate(ID = dplyr::row_number())

projected_model <- as.matrix(scaled_pbmc_data_model) %*% projection_scaled

model_df <- dplyr::left_join(
  distance_df_small_edges_pbmc |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

axes_obj <- gen_axes(
  proj = projection,
  limits = 1.35,
  axis_pos_x = -0.75,
  axis_pos_y = -0.75,
  axis_labels = names(scaled_pbmc_data),
  threshold = 0.12)

@@

data_mnist <- training_data_mnist |>
  select(-ID) |>
  mutate(type = "data")

mnist_scaled_obj <- gen_scaled_data(
  data = tsne_mnist2)
tsne_minst_scaled <- mnist_scaled_obj$scaled_nldr 
```

@@

<!-- Fit the model and compute error-->
```{r}
#| label: hexbin-mnist
## Compute hexbin parameters
num_bins_x_mnist <- 30
lim1 <- mnist_scaled_obj$lim1
lim2 <- mnist_scaled_obj$lim2
r2_mnist <- diff(lim2)/diff(lim1)

mnist_model <- fit_highd_model(
  training_data = training_data_mnist,
  emb_df = tsne_minst_scaled,
  bin1 = num_bins_x_mnist,
  r2 = r2_mnist,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = TRUE,
  col_start_highd = "PC"
)

df_bin_centroids_mnist <- mnist_model$df_bin_centroids
df_bin_mnist <- mnist_model$df_bin

## Triangulate bin centroids
tr1_object_mnist <- tri_bin_centroids(
  df_bin_centroids_mnist, x = "c_x", y = "c_y")
tr_from_to_df_mnist <- gen_edges(
  tri_object = tr1_object_mnist)

## Compute 2D distances
distance_mnist <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_mnist, 
  start_x = "x_from", 
  start_y = "y_from", 
  end_x = "x_to", 
  end_y = "y_to", 
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_mnist <- find_lg_benchmark(
  distance_edges = distance_mnist, 
  distance_col = "distance")

@@

tr_df <- distinct(tibble::tibble(
  x = c(tr_from_to_df_mnist[["x_from"]], tr_from_to_df_mnist[["x_to"]]), 
  y = c(tr_from_to_df_mnist[["y_from"]], tr_from_to_df_mnist[["y_to"]])))

@@

tr_from_to_df_mnist <- inner_join(
  tr_from_to_df_mnist, distance_df_small_edges_mnist, 
  by = c("from", "to")) |>
  mutate(ID = row_number())

tsne_minst_scaled_temp <- tsne_minst_scaled |>
  mutate(cluster = if_else((tSNE1 <= 1) & (tSNE1 >= 0.78) & (tSNE2 <= 0.62) & (tSNE2 >= 0.39), "small_clust", "big_clust"))

trimesh_removed_mnist <- ggplot() + 
  geom_segment(data = tr_from_to_df_mnist, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_minst_scaled_temp,
             aes(
               x = tSNE1,
               y = tSNE2,
               colour = cluster
             ),
             alpha=0.2)+
  scale_color_manual(values=c("#999999",'#ff7f00')) +
  interior_annotation("a", c(0.08, 0.93), cex = 1.5) +
  theme(legend.position = "none")

tsne_plot_mnist_clust <- ggplot() + 
  geom_point(data = tsne_minst_scaled_temp,
             aes(
               x = tSNE1,
               y = tSNE2,
               colour = cluster
             ),
             alpha=0.2)+
  scale_color_manual(values=c('#d9d9d9', '#ff7f00')) +
  interior_annotation("a", c(0.08, 0.93), cex = 2) +
  theme(legend.position = "none")

@@

projected_df <- proj_obj1$projected_df |>
  mutate(cluster = tsne_minst_scaled_temp$cluster)

@@

projected_df <- proj_obj2$projected_df |>
  mutate(cluster = tsne_minst_scaled_temp$cluster)

>>>>

```{r}
#| label: gen-vis-obj

gen_nldr_vis_algo_obj <- function(high_d_data, nldr_data, num_x_bins, col_start = "x") {

  nldr_scaled_obj <- gen_scaled_data(
    data = nldr_data)

  nldr_scaled <- nldr_scaled_obj$scaled_nldr |>
    mutate(ID = row_number())

  lim1 <- nldr_scaled_obj$lim1
  lim2 <- nldr_scaled_obj$lim2
  r2 <- diff(lim2)/diff(lim1)

  ## hexagon binning to have regular hexagons
  hb_obj <- hex_binning(
    data = nldr_scaled, 
    bin1 = num_x_bins, 
    r2 = r2,
    q = 0.1)

  ## Data set with all centroids
  all_centroids_df <- hb_obj$centroids

  a1_1 <- hb_obj$a1

  ## Generate all coordinates of hexagons
  hex_grid <- hb_obj$hex_poly

  ## To obtain the standardise counts within hexbins
  counts_df <- hb_obj$std_cts
  df_bin_centroids <- extract_hexbin_centroids(
    centroids_df = all_centroids_df, 
    counts_df = counts_df) |>
    filter(drop_empty == FALSE)

  nldr_with_hb_id <- hb_obj$data_hb_id
  df_all <- dplyr::bind_cols(high_d_data, 
                                       nldr_with_hb_id)
  df_bin <- avg_highd_data(data = df_all, col_start = col_start)

  hex_grid_nonempty <- hex_grid |>
    filter(hex_poly_id %in% df_bin_centroids$hexID)

  tr1_object <- tri_bin_centroids(
    df_bin_centroids, x = "c_x", y = "c_y")
  tr_from_to_df <- gen_edges(
    tri_object = tr1_object)

  ## Compute 2D distances
  distance_df <- cal_2d_dist(
    tr_coord_df = tr_from_to_df, 
    start_x = "x_from", 
    start_y = "y_from", 
    end_x = "x_to", 
    end_y = "y_to", 
    select_vars = c("from", "to", "distance"))

  ## To find the benchmark value
  benchmark <- find_lg_benchmark(
    distance_edges = distance_df, 
    distance_col = "distance")

  return(list(nldr_scaled = nldr_scaled,
              hex_grid_with_counts = hex_grid_with_counts,
              hex_grid_nonempty = hex_grid_nonempty,
              df_bin_centroids = df_bin_centroids,
              tr_from_to_df = tr_from_to_df,
              df_bin = df_bin,
              distance_df = distance_df,
              benchmark = benchmark))

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
#| label: hexbin-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 60

algo_obj_pbmc <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_pbmc, 
  nldr_data = umap_pbmc, 
  num_x_bins = num_bins_x_pbmc,
  col_start = "pc")

umap_pbmc_scaled <- algo_obj_pbmc$nldr_scaled
distance_pbmc <- algo_obj_pbmc$distance_df
tr_from_to_df_pbmc <- algo_obj_pbmc$tr_from_to_df
benchmark_pbmc <- algo_obj_pbmc$benchmark
df_bin_centroids_pbmc <- algo_obj_pbmc$df_bin_centroids
df_bin_pbmc <- algo_obj_pbmc$df_bin

@@

# # Combine with the true model for visualization
# df <- dplyr::bind_rows(scaled_pbmc_data_model |> mutate(type = "model"),
#                        scaled_pbmc_data |> mutate(type = "data"))

@@

<!--Fit the best model and compute error-->
```{r}
#| label: num-bins-pbmc

## Compute hexbin parameters
num_bins_x_pbmc <- 30

algo_obj_pbmc <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_pbmc, 
  nldr_data = umap_pbmc2, 
  num_x_bins = num_bins_x_pbmc,
  col_start = "pc")

umap_pbmc_scaled_best <- algo_obj_pbmc$nldr_scaled
distance_pbmc <- algo_obj_pbmc$distance_df
tr_from_to_df_pbmc <- algo_obj_pbmc$tr_from_to_df
benchmark_pbmc <- algo_obj_pbmc$benchmark
df_bin_centroids_pbmc <- algo_obj_pbmc$df_bin_centroids
df_bin_pbmc <- algo_obj_pbmc$df_bin

@@

trimesh_removed_pbmc_best <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_pbmc_scaled_best,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.3,
    color = "#999999"
  )  +
  interior_annotation("b1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

@@

@@

@@

@@

data_mnist <- training_data_mnist |>
  select(-ID) |>
  mutate(type = "data")
```

@@

<!-- Fit the model and compute error-->
```{r}
#| label: hexbin-mnist

num_bins_x_mnist <- 30

algo_obj_mnist <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_mnist, 
  nldr_data = tsne_mnist2, 
  num_x_bins = num_bins_x_mnist,
  col_start = "PC")

tsne_minst_scaled <- algo_obj_mnist$nldr_scaled
distance_mnist <- algo_obj_mnist$distance_df
tr_from_to_df_mnist <- algo_obj_mnist$tr_from_to_df
benchmark_mnist <- algo_obj_mnist$benchmark
df_bin_centroids_mnist <- algo_obj_mnist$df_bin_centroids
df_bin_mnist <- algo_obj_mnist$df_bin

@@

@@

tsne_minst_scaled <- tsne_minst_scaled |>
  mutate(cluster = if_else((tSNE1 <= 1) & (tSNE1 >= 0.78) & (tSNE2 <= 0.62) & (tSNE2 >= 0.39), "small_clust", "big_clust"))

trimesh_removed_mnist <- ggplot() + 
  geom_segment(data = tr_from_to_df_mnist, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_minst_scaled,
             aes(
               x = tSNE1,
               y = tSNE2,
               colour = cluster
             ),
             alpha=0.2)+
  scale_color_manual(values=c("#999999",'#ff7f00')) +
  interior_annotation("a", c(0.08, 0.93), cex = 1.5) +
  theme(legend.position = "none")

tsne_plot_mnist_clust <- ggplot() + 
  geom_point(data = tsne_minst_scaled,
             aes(
               x = tSNE1,
               y = tSNE2,
               colour = cluster
             ),
             alpha=0.2)+
  scale_color_manual(values=c('#d9d9d9', '#ff7f00')) +
  interior_annotation("a", c(0.08, 0.93), cex = 2) +
  theme(legend.position = "none")

@@

projected_df <- proj_obj1$projected_df |>
  mutate(cluster = tsne_minst_scaled$cluster)

@@

projected_df <- proj_obj2$projected_df |>
  mutate(cluster = tsne_minst_scaled$cluster)
